Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-08-14 analysis/paper/paper.Rmd
IMPLEMENT {targets} PIPELINE; UPDATE DATA ETL w/ NEW ADFG RECORDS
major revamp of compendium and code inspired by updated records from ADFG. implemented the {targets} package workflow.

```{r setup, include = FALSE}
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

```{r getdata}
con <- dbConnect(
  odbc::odbc(),
  dsn = "PostgreSQL pep",
  uid = keyringr::get_kc_account("pgpep_londonj"),
  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
  sslmode = "disable"
)

# qry removes two Sea of Okhotsk bearded seals
qry <- {
  "SELECT * 
  FROM telem.res_iceseal_haulout_cov 
  WHERE 
  EXTRACT(MONTH FROM haulout_dt) IN (3,4,5,6,7) AND
  rast_vwnd IS NOT NULL AND
  species != 'Ph' AND
  speno != 'EB2009_7010' AND
  speno != 'EB2009_3002' "
}

dat.sf <- sf::st_read(con,query = qry) %>% 
  filter(lubridate::yday(haulout_dt) < 197 ) %>% 
  mutate(age = case_when(
    age %in% c("pup", "PUP") ~ "YOUNG OF YEAR",
    age %in% c("yearling","YEARLING","subadult","Subadult") ~ "SUBADULT",
    age %in% c("adult","ADULT","Adult") ~ "ADULT",
    TRUE ~ age),
    species = case_when(
    species == "bearded" ~ "Bearded seal",
    species == "spotted" ~ "Spotted seal",
    TRUE ~ species),
    sex = case_when(
    sex %in% c("Female","female") ~ "F",
    sex %in% c("Male","male") ~ "M",
    TRUE ~ sex)
    ) %>% 
  filter(!is.na(age))

load(here::here('data/locs_sf.Rdata'))

deploy_meta <- locs_sf %>% 
  as_data_frame() %>% 
  select(speno, deploy_dt) %>% 
  filter(speno %in% !!unique(dat.sf$speno)) %>% 
  group_by(speno) %>% 
  summarise(deploy_dt = min(deploy_dt, na.rm = TRUE))

dat.sf <- dat.sf %>% left_join(deploy_meta) %>% 
  mutate(bday = as.Date(paste0(lubridate::year(deploy_dt),"-03-01")),
         age_days = difftime(haulout_dt, bday, tz="UTC", units = "days"),
         age = case_when(
           age_days > 360 & age == "YOUNG OF YEAR" ~ "SUBADULT",
           TRUE ~ age
         )) %>% 
  mutate(age = forcats::fct_relevel(age, c("ADULT", "SUBADULT", "YOUNG OF YEAR"))) %>% 
  dplyr::select(-c(bday,age_days)) %>% 
  relocate(deploy_dt, .after = age) %>% 
  filter(!(species == "Bearded seal" & age == "YOUNG OF YEAR"))

# read in world land polygons and buffer in by 100m; want to ensure only locations well
world <- ne_countries(scale = "large", returnclass = "sf") %>% 
  filter(continent != "Antarctica") %>% 
  st_wrap_dateline() %>% st_transform(st_crs(dat.sf)) %>% 
  st_buffer(-100)

not_intersects <- function(x,y) {
  !sf::st_intersects(x,y)
}

dat.sf <- dat.sf %>% 
  sf::st_filter(st_combine(world), .predicate = not_intersects)

save("dat.sf", file = here::here("data/dat_sf.Rdata"))

grid.sf <- sf::st_read(con,
                       query = "SELECT * FROM base.geo_analysis_grid") %>% 
  st_filter(st_bbox(dat.sf) %>% st_as_sfc())

dbDisconnect(con)

deploy_tbl <- dat.sf %>% 
  tibble::as_tibble() %>% 
  dplyr::group_by(speno,species, sex, age) %>% 
  dplyr::summarise(min_dt = min(haulout_dt),
            max_dt = max(haulout_dt),
            n_days = difftime(max_dt, min_dt, units = "days"),
            n_hours = n()
  ) %>% 
  dplyr::select(-c("min_dt","max_dt"))

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was filtered to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class (non-dependent *young-of-the-year*, sexually
immature *subadults*, and mature *adults*) were determined at the time of deployment by
subjective judgement based on length, and pelage characteristics for some species. Field
identification of age class can be inexact, particularly when discerning subadults from
adults. In the case of ribbon seals, pelage differences between subadults and adults are
distinguishable, with subadults often having less defined ribbons. Bearded seal subadults
will often have a spotted pattern in the pelage that is not seen in adults. Spotted seal
pelage is monomorphic and can not be used to discern age class. Despite these challenges,
we feel the age classifications used in this analysis are useful categories in testing for
age-related effects on haul-out behavior. Seals determined to be less than one year were
classified as young-of-the-year. Subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

deploy_tbl %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

```{r seal-hours}
seal_hours_tbl <- dat.sf %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
                  forcats::fct_relevel("Sun","Mon","Tue","Wed","Thu","Fri","Sat")
  ) %>% 
  dplyr::group_by(species, wkofyr, dayofwk) %>% 
  dplyr::count()
```

@@

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May. Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt)) %>% 
  filter(month != 'Jun')

@@

```{r locs-count-mapdata}
haulout_locs <- dat.sf %>% 
  filter(lubridate::hour(haulout_dt) == 12)

>>>>

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

```{r getdata}
targets::tar_load(analysis_data, store = here::here("_targets"))
targets::tar_load(grid, store = here::here("_targets"))
targets::tar_load(deploy_table, store = here::here("_targets"))
```

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was filtered to include only records from 1 March to
15 July between `r min(lubridate::year(analysis_data$haulout_dt))` and
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class (non-dependent *young-of-the-year*, sexually
immature *subadults*, and mature *adults*) were determined at the time of deployment by
subjective judgement based on length, and pelage characteristics for some species. Field
identification of age class can be inexact, particularly when discerning subadults from
adults. In the case of ribbon seals, pelage differences between subadults and adults are
distinguishable, with subadults often having less defined ribbons. Bearded seal subadults
will often have a spotted pattern in the pelage that is not seen in adults. Spotted seal
pelage is monomorphic and can not be used to discern age class. Despite these challenges,
we feel the age classifications used in this analysis are useful categories in testing for
age-related effects on haul-out behavior. Seals determined to be less than one year were
classified as young-of-the-year. Subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

deploy_table %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

```{r seal-hours}
seal_hours_tbl <- analysis_data %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
                  forcats::fct_relevel("Sun","Mon","Tue","Wed","Thu","Fri","Sat")
  ) %>% 
  dplyr::group_by(species, wkofyr, dayofwk) %>% 
  dplyr::count()
```

@@

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May. Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
d <- analysis_data %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt)) %>% 
  filter(month != 'Jun')

@@

```{r locs-count-mapdata}
haulout_locs <- analysis_data %>% 
  filter(lubridate::hour(haulout_dt) == 12)
